Manuscript_draft

### 2020-12-12 manuscript_draft.Rmd
start final manuscript process

@@

# Let's filter only for trees that have measurements in 1992 and 2019 
treeList<- # assign variable
  X %>% 
  filter(TreeID!="New") %>% #remove new trees
  mutate(TreeID=paste(Plot,TreeID,sep="-")) %>% # create unique treeID
  filter(Species %in% c("Spruce","Fir")) %>% # filter birch and fdi out of dataset
  filter(Status=="Live") %>% # filter for live trees
  filter(Year %in% c(1992,2019)) %>% #filter for start and end dates
  group_by(TreeID) %>% 
  summarise(N=n()) %>% 
  filter(N==2) %>% #filter for trees with 5 measurements
  pull(TreeID)

# derive periodic increment (period diameter or basal area increment)
piDat<-
  dat %>% 
  mutate(TreeID=paste(Plot,TreeID,sep="-")) %>% # create unique treeID
  filter(TreeID %in% treeList) %>% 
  mutate(BA=d2ba(DBH)) %>% # convert DBH to basal area
  dplyr::select(Plot,TreeID,Species,Year,BA) %>% 
  pivot_wider(names_from=Year,values_from=BA) %>% 
  mutate(PI=`2019`-`1992`) %>% # 
  filter(PI>0) %>% # one tree has a PI of 0, has to be removed
  mutate(PIpct=PI/`1992`*100) %>% # express periodic increment a a percentage
  dplyr::select(Plot,TreeID,Species,PBAI=PI,PBAIpct=PIpct,Init=`1992`) %>% 
  left_join(plotDat,by="Plot")

@@

  fitBA.summ<-
    list(table=tab_model(fitBA),
         caption="Summary of mixed effects model of tree-level growth")

@@

  fitBApct.summ<-
    list(table=tab_model(fitBApct),
         caption="Summary of mixed effects model of tree-level periodic basal area % increment (PBAI %")

@@

treeBAI.plot<-

  emmeans(rg,~Init|BA.Target+Species,type="response") %>% 
  as.data.frame() %>%
  mutate(Init=ba2d(Init)) %>% # convert initial basal area back to diameter

@@

# set variable
sppContrast<-
  list(caption="Contrasts between predicted fir and spruce growth across treatment units and initial diameter classes.  Ratio between predicted fir and growth is in 'ratio' column.",

  table=

  pairs(rg,by=c("BA.Target","Init"),type="response") %>% 
  as.data.frame() %>%
    mutate(Init=as.numeric(as.character(Init))) %>% # change from factor to numeric
    mutate(Init=ba2d(Init)) %>% # convert basal area to diameter
  arrange(BA.Target,Init) %>% 
  rename(DBH.1992="Init") %>% 
  mutate_if(is.double,round,5) %>%  # round to five digits

  # format table

  # flextable
  flextable(col_keys = c(colnames(.)[c(1:2,6:8,3:5)])) %>%

  align(align="left",part = "all") %>% 
  vline( border = fp_border(color="#2c3a51" , width = 1), part = "all" ) %>% 
  hline( border = fp_border(color="#2c3a51" , width = 1), part = "all" ) %>% 
  merge_v(j=c(1:2,6:8)) %>% 
  valign(valign="top") %>%
 border_outer(part="all", border = fp_border(color="#2c3a51" , width = 2)) %>% 
   fix_border_issues()  # border(border.bottom = fp_border(color="black"))
) # close list

@@

The tree-level basal area increment model indicated that interactions between log-transformed tree size, species and treatment explained `r round(r.squaredGLMM(fitBA)[1]*100,1)`% (marginal R²) of the variation in tree log-transformed 1992-2019 BAI (`r tableNums("fitBA.table",display="cite")`).  Estimated marginal means from the model showed that between species and among treatments, larger trees increased basal area more than smaller trees (`r figNums("treeGrowth.plot",display="cite")`).  Species contrasts showed that spruce trees increased  basal area more than than fir in the high-removal treatment unit, whereas basal growth between the two species was close to equal in the low-removal and control units (`r appNums("sppContrast.table",display="cite")`).  Within each species, treatment contrasts showed that spruce basal area increment in the high-removal unit exceeded spruce growth in both other units (`r appNums("trtContrast.table",display="cite")`), whereas fir basal area increment in the high- and low-removal units was higher than the control unit, and there were no significant differences between the two harvested units (`r appNums("trtContrast.table",display="cite")`).

@@

### Mortality

```{r mortality graphs}

# Create dead dataset

# Dead species totals
deadList<-
  X %>% 
  filter(Status=="Dead") %>% 
  mutate(ID=paste(Plot,TreeID,sep="-")) %>% 
  group_by(ID) %>% 
  summarise_all(first)

dead<-
  X %>% 
  filter(Status=="Dead") %>% 
  mutate(ID=paste(Plot,TreeID,sep="-")) %>% 
  mutate(Species=fct_recode(Species,`Douglas-fir`="Fd")) %>% 
  group_by(ID) %>% 
  summarise_all(first) %>% 
  ungroup() %>% 
  mutate(BA=d2ba(DBH)) %>% # convert to basal area per hectare 
  group_by(BA.Target,DBH.class,Year) %>% 
  summarise(BA=sum(BA),
            k=n())

dead.Summ<-
  dead %>% 
  ungroup() %>% 
  group_by(BA.Target) %>% 
  summarise(BA=sum(BA)/sum(dead$BA)*100,
            k=sum(k)/sum(dead$k)*100) %>% 
  mutate_if(is.numeric,round,1)

p1<-
  dead %>% 
  ggplot(aes(x=DBH.class,y=k,fill=Year))+
  geom_bar(stat="identity")+
  ylab("Total tree mortality (stems)")+
  facet_wrap(~BA.Target)+
  xlab("")+
  theme(axis.text.x = element_blank())

p2<-
  dead %>% 
  ggplot(aes(x=DBH.class,y=BA,fill=Year))+
  geom_bar(stat="identity")+
  ylab("Total basal area mortality (m²)")+
  facet_wrap(~BA.Target)

mort.plot<-cowplot::plot_grid(p1, p2, ncol = 1, align = "v")
figNums("mort.plot",display=FALSE)

A total of `r deadList %>% nrow()` trees (`r deadList %>% filter(Species=="Fir") %>% nrow()` fir, `r deadList %>% filter(Species=="Spruce") %>% nrow()` spruce, and `r deadList %>% filter(Species=="Fd") %>% nrow()` Douglas-fir) died from 1992 to 2019.  No tree mortality was noted during the 1992 and 1994 data collection.  By treatment unit, the control unit had the highest mortality, in terms of trees and basal area (
`r dead.Summ %>% filter(BA.Target=="TU:Control") %>% pull("BA")` and 
`r dead.Summ %>% filter(BA.Target=="TU:Control") %>% pull("BA")`% of tree mortality, respectively).  The high-removal treatment unit had the lowest mortality by number of trees and basal area (
`r dead.Summ %>% filter(BA.Target=="TU:10") %>% pull("BA")` and 
`r dead.Summ %>% filter(BA.Target=="TU:10") %>% pull("BA")`% of tree mortality, respectively).  Tree mortality was concentrated in the smaller diameter classes in the control and low-removal treatment unit, and most mortality occured during the most recent measurement period of 2009-2019 (
`r figNums("mort.plot",display="cite")`).

```{r display mortality plot}
mort.plot
```

`r figNums("mort.plot",caption="Total tree mortality recorded in plots from 1992-2019, presented as total stems (top panel) and basal area (bottom panel).")`

From Mike's email:
Eventually as BA stand density increases, density dependent mortality kicks in, because there is only so much growing space or access to site resources to go around. This can occur with the stand (conceptually or hypothetically)  in two possible ways: (a) “classic” suppression of smaller tree classes through asymmetric competition of bigger trees shading out smaller ones, or (b) high stand densities suppressing the vigor of the whole stand through various means, suggesting more symmetric competition – i.e. the idea that small trees compete for growing space with big ones as well. Asymmetric competition is generally accepted by most foresters and silviculturists; symmetric competition less so, and is often harder to demonstrate.

*For the discussion: it seems that we're seeing mortality increasing in the plots, especially the control unit over the last 11 years.  Does this reflect the density-dependent mortality that Mike refers to above?.*

More from Mike on mortality:

The thing about unevenaged or complex stand management, esp in ‘wilder’ BC stands, is that we are working in stands with a long legacy of pre-existing ecosystem processes before we enter the stand to manipulate them. Clearcutting liquidates these legacies for the most part, and traditional silviculture and GY has not had to deal with mortality much, other than clearly density dependent suppression of smaller dbh classes. [Side issue: most BC experience and data with unevenaged stands has been in IDF, which have been fire excluded on dry sites, and unevenaged structures allowed to develop. We must be careful about comparing Summit data to IDF precedents, because of substantial differences in ecology.]

And we are working a more extensive level of management in the SBS and ESSF,  where our stands are older, and not manicured nor vacuumed up for mortality and blowdown. From an ecosystem management point of view, that Is probably a good thing.

So warts and all, I think mortality is an inherent characteristic of the types of spruce-balsam stands and management we are studying. Expecting and documenting mortality was built into the EP 1162 working plan. The treatment effects we anticipate are the net stand growth = gross growth minus mortality.

Conceptually, tree mortality following the 1991/92 harvest (and subsequent sanitation felling)  functions as a BA removal from the live-tree pool. In most plots, BA of trees that die within a measurement period is subsumed or overridden by the total BA growth of the rest of the trees during that period.

Where the periodic data trends get messier is where single big trees or a cluster of trees gets killed by something, resulting in a pulse of BA mortality. Examples: The great big Fdi in Plot 18, or notably, about 3-4 big Sx in Plot 3 that got nailed by spruce beetle.

In our results and discussion:
1.	We can report on what stand growth and development trends seem to be clearly and unambiguously  influenced by initial BA conditions, and the magnitude of these treatment effects. 
2.	A component of tree mortality (e.g. certain size classes) may be density dependent. 
3.	A component of tree mortality may be density-independent (a.k.a stochastic or chaotic). 
4.	By reporting on all the above, we are providing managers with results that integrate silvicultural expectations observations from real-life “messy” stands which will experience all of the above. By keeping our eye on overall stand growth and performance following treatment, we can inform / reassure managers that individual tree mortality events are not the end of the world, but rather need to be viewed in the context of overall stand dynamics and growth expectations.

@@

>>>>

---
title: "Manuscript_draft"
author: "Michael Jull and Hardy Griesbauer"
date: "10/12/2020"
output:
  word_document:
    reference_docx: wordTemplate2.docx
  html_document:
    df_print: paged
csl: cjfr.csl
bibliography: ep1162.bib

---

```{r load libraries for analyses, echo=FALSE, message=FALSE, warning=FALSE}

knitr::opts_chunk$set(echo = FALSE,message=FALSE,warning=FALSE)

# Load libraries
library(tidyverse)
library(lme4)
library(lmerTest)
library(broom)
library(MuMIn)
library(ggeffects)
library(car)
library(flextable)
library(captioner)
library(lsmeans)
library(cowplot)
library(officer)
library(sjPlot)
library(emmeans)

load("data/ep1162_Data.RData")
load("data/ep1162_plotData.RData")

# SOME PRELIM DATA FORMATTING
# formatting plotDat
plotDat<-
  plotDat %>% 
  mutate(Plot=as.factor(Plot))

# set up diameter classes
diamClass=seq(from=7.5,to=100,by=5)

# Set against scientific notation
options(scipen=999)

# Merge data
X<-
  dat %>% 
  left_join(plotDat,by="Plot") %>% 
  mutate(DBH.class=cut(DBH,diamClass,include.lowest=TRUE,ordered_result = TRUE,right=F,
                       labels=diamClass[1:(length(diamClass)-1)])) %>% 
  mutate(Plot=as.factor(Plot)) %>%

  # replace species codes with words
  mutate(Species=recode_factor(Species,Bl="Fir",Sx="Spruce")) %>% 
  mutate(Species=fct_drop(Species)) %>% # have an extra level in here for some reason

  # Create TU lookup
  mutate(BA.Target=paste0("TU:",BA.Target))

# set up captions

# Code to create captions for tables
tableNums <- captioner(prefix = "Table")

# Code to create captions for figures
figNums <- captioner(prefix="Figure")

# Code to create captions for appendices
appNums <- captioner(prefix="Appendix")

# Set ggplot theme
theme_set(theme_bw())

# Define function to convert diameter (cm) to basal area (m2), and vice versa
d2ba<-function(diam) {return((diam/2/100)^2*pi)}
ba2d<-function(ba) {return(sqrt(ba/pi)*200)}

This section will describe the methods we used for data analyses, not field data collection.

From our technical report submitted to FCI earlier:

Our data analyses and summaries examined the effects of the different levels of stand-level basal area density treatments on three main attributes of stand development following treatment. These include, in order:

1. Post-treatment stand basal area dynamics, including the rate and pattern of basal area re-growth and recovery over time following different levels of initial basal area density reductions;
2. Stand structural outcomes, including treatment influences on the overall abundance of trees (in sph) by diameter class, within the different treatment types, and;
3. Tree species composition outcomes, including treatment influences on the relative abundance and size class distributions of subalpine fir and hybrid white spruce within the different treatment types.

Analyses are grouped into three main categories (below).  These are treated as separate categories in the Results section.

1. Tree-level responses to treatment;
2. Stand-level responses to treatment:

```{r define captions to be used in methods results and discussion, echo=FALSE}

# Captions (put these in order)

  # Introductory 
  appNums(name="paiTable",caption="Number of growing seasons between each measurement period",display=FALSE) # Table showing number of growing seasons between each measurement period
  appNums(name="standVol.plot",caption="Plot total volume for five measurement periods",display=FALSE) # Plot volume figure
  appNums(name="standBA.plot",caption="Plot basal area for five measurement periods",display=FALSE)  # Plot basal area figure
  appNums(name="standQMD.plot",caption="Plot quadratic mean diameter for five measurement periods",display=FALSE) # QMD figure  
  appNums(name="standK.plot",caption="Plot tree density for five measurement periods", display=FALSE) # SPH figure

  # Tree heights model
  appNums("heightDiamRawPlot",display=FALSE) # Plot of diameter/heights`

  # Diameter class analysis
  figNums("diamVol.plot", caption="Mean plot volume distribution by diameter class and treatment, for two periods.  Black line shows total volume.  Volume calculated on trees meeting 17.5cm DBH limit.  Spruce and fir contribution to volume by diameter class are shown with bar plots.  Douglas-fir volume contribution by diameter class is not shown.",display="FALSE")

  figNums("diamBA.plot", caption="Mean plot basal area distribution by diameter class and treatment, for two periods.  Black line shows total basal area.  Spruce and fir contribution by diameter class are shown ith bar plots.  Douglas-fir contribution by diameter class is not shown.",display="FALSE")

  figNums("diamK.plot", caption="Mean tree density distribution by diameter class and treatment, for two periods.  Black line shows total tree density.  Spruce and fir contribution by diameter class are shown ith bar plots.  Douglas-fir contribution by diameter class is not shown.",display="FALSE")

## Tree-level growth responses

To test the hypothesis that tree-level radial growth response over the 1992-2019 period varied with interactions between (i) tree diameter at the time of treatment (1992); (ii) treatment; and (iii) species, we fit linear mixed effects models with log-transformed 1992-2019 diameter increment as the response variable, and treatment, species and log-transformed 1992 diameter as interacting fixed effects, and plot as a random intercept effect.  We also fit a similar model with the 1992-2019 diameter increment expressed as a percentage of 1992 diameter.  Plots of residuals and fit lines with orginal data were examined to evaluate model goodness of fit.  Variance explained for each model are reported using the approach described in Nakagawa et al. [-@Nakagawa2013], as implemented in the MuMIn package for R [@add later].

## Stand-level growth and structure

### Stand-level volume and basal area estimates
Total and merchantable volumes were estimated for trees using species- and region-specific equations [@nigh2016] based on tree height and DBH.  Consistent with forest management practices in British Columbia, we estimated volumes for trees meeting a minimum diameter utilization limit of 17.5cm DBH.  For trees lacking a tree height measurement, we estimated height by fitting a single diameter-height model developed with a linear mixed effects model, with log-transformed DBH as a fixed effect, and tree nested within plot as a random effect.  Trees with broken tops were omitted from diameter-height models. We summed tree-level estimates of volume and basal area to generate stand-level estimates for each plot. To test the hypothesis that changes in stand attributes over 27 years varied by treatment, we fit separate linear models with stand total volume, basal area, and density as response variables, and the interaction between treatment and year as the independent variable.

@@

# Filter only for trees that have measurements in 1992 and 2019 
treeList<- # assign variable
  X %>% 
  filter(TreeID!="New") %>% # remove new trees
  mutate(TreeID=paste(Plot,TreeID,sep="-")) %>% # create unique treeID
  filter(Species %in% c("Spruce","Fir")) %>% # filter birch and fdi out of dataset
  filter(Status=="Live") %>% # filter for live trees
  filter(Year %in% c(1992,2019)) %>% #filter for start and end dates
  group_by(TreeID) %>% 
  summarise(N=n()) %>% 
  filter(N==2) %>% #filter for trees with 5 measurements
  pull(TreeID)

# derive periodic increment (period diameter or basal area increment)
piDat<-
  X %>% 
  mutate(TreeID=paste(Plot,TreeID,sep="-")) %>% # create unique treeID
  filter(TreeID %in% treeList) %>% 
  mutate(BA=d2ba(DBH)) %>% # convert DBH to basal area
  dplyr::select(Plot,TreeID,Species,Year,BA) %>% 
  pivot_wider(names_from=Year,values_from=BA) %>% 
  mutate(PI=`2019`-`1992`) %>% # 
  filter(PI>0) %>% # one tree has a PI of 0, has to be removed
  mutate(PIpct=PI/`1992`*100) %>% # express periodic increment a a percentage
  dplyr::select(Plot,TreeID,Species,PBAI=PI,PBAIpct=PIpct,Init=`1992`) %>% 
  left_join(plotDat,by="Plot")

@@

@@

@@

@@

sppContrast<-

  pairs(rg,by=c("BA.Target","Init"),type="response") %>% 
  as.data.frame() %>%
    mutate(Init=as.numeric(as.character(Init))) %>% # change from factor to numeric
    mutate(Init=ba2d(Init)) %>% # convert basal area to diameter
  arrange(BA.Target,Init) %>% 
  rename(DBH.1992="Init")

@@

```{r set up tree BAI references}

  figNums(name="treeBAI.plot",display=FALSE)

  tableNums("fitBA.table",display=FALSE) # Summary of mixed effects model for BAI
  tableNums("fitBApct.table",display=FALSE) # Summary of mixed effects model for BAI

  appNums("sppContrast.table",display=FALSE) # Species contrasts from treeBAI model
  appNums("trtContrast.table",display=FALSE) # Treatment contracts from treeBAI model

The tree-level basal area increment model indicated that interactions between log-transformed tree size, species and treatment explained `r round(r.squaredGLMM(fitBA)[1]*100,1)`% (marginal R²) of the variation in log-transformed 1992-2019 tree-level BAI (`r tableNums("fitBA.table",display="cite")`).  Estimated marginal means from the model showed that between species and among treatments, larger trees increased basal area more than smaller trees (`r figNums("treeGrowth.plot",display="cite")`).  Species contrasts showed that spruce trees increased  basal area more than than fir in the high-removal treatment unit, whereas basal growth between the two species was close to equal in the low-removal and control units (`r appNums("sppContrast.table",display="cite")`).  Within each species, treatment contrasts showed that spruce basal area increment in the high-removal unit exceeded spruce growth in both other units (`r appNums("trtContrast.table",display="cite")`), whereas fir basal area increment in the high- and low-removal units was higher than the control unit, and there were no significant differences between the two harvested units (`r appNums("trtContrast.table",display="cite")`).

@@

@@

\newpage

# Tables

`r tablesNums("fitBA.table",caption="Summary of mixed effects model of tree-level basal area increment")`

```{r summarize tree level BAI model}

\newpage

`r tablesNums("fitBApct.table",caption="Summary of mixed effects model of tree-level basal area increment as percentage of initial tree size")`

```{r summarize tree level BAI as percentage model}

\newpage

# Figures

# Using estimated marginal means
  emmeans(rg,~Init|BA.Target+Species,type="response") %>% 
  as.data.frame() %>%
  mutate(Init=ba2d(Init)) %>% # convert initial basal area back to diameter

\newpage

# Appendices

`r appNums("paiTable",caption="Number of growing seasons between measurement periods")`

```{r show paiTable, ft.align="left"}

  tibble(`Measurement Year`=c(1992,1994,1997,2009,2019),`Growing Seasons`=c(NA,3,3,11,11)) %>% 
  flextable() %>% 
  align(align="left",part="all") %>% 
  # set_caption(caption="Number of growing seasons between measurement periods") %>% 
  autofit()

\newpage

`r appNums("sppContrast.table",caption="Contrasts between predicted fir and spruce growth across treatment units and initial diameter classes.  Ratio between predicted fir and growth is in 'ratio' column.")`

```{r treelevel species contrast table}

# Tree-level species contrast table
  sppContrast %>%

  mutate_if(is.double,round,5) %>%  # round to five digits

  # format table
  flextable(col_keys = c(colnames(.)[c(1:2,6:8,3:5)])) %>% 
  align(align="left",part = "all") %>% 
  vline( border = fp_border(color="#2c3a51" , width = 1), part = "all" ) %>% 
  hline( border = fp_border(color="#2c3a51" , width = 1), part = "all" ) %>% 
  merge_v(j=c(1:2,6:8)) %>% 
  valign(valign="top") %>%
 border_outer(part="all", border = fp_border(color="#2c3a51" , width = 2)) %>% 
   fix_border_issues()  # border(border.bottom = fp_border(color="black"))

\newpage

`r appNums("trtContrast.table",caption="Contrasts of predicted 1992-2019 tree growth among treatments within each species.  Ratio of predicted growth between treatments is in 'ratio' column.")`

```{r tree level treatment contrasts}

# Treatment contrast

trtContrast %>%

  mutate_if(is.double,round,5) %>%

  # flextable
  flextable(col_keys = c(colnames(.)[c(2,1,6:8,3:5)])) %>% 
  border_outer(part="all", border = fp_border(color="#2c3a51" , width = 2)) %>% 
  align(align="left",part = "all") %>% 
  vline( border = fp_border(color="#2c3a51" , width = 1), part = "all" ) %>% 
  hline( border = fp_border(color="#2c3a51" , width = 1), part = "all" ) %>% 
  merge_v(j=c(1:2,6:8)) %>% 
  valign(valign="top") %>%
  hline_bottom() %>% 
  fix_border_issues() # border(border.bottom = fp_border(color="black"))

\newpage
